---
title: "BMTR spread kernel: choosing gamma/kl/s1"
date: "2026-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
devtools::load_all("..", quiet = TRUE)
library(ggplot2)
```

## What this is

A record of how the spread kernel hyperparameter defaults (`gamma`, `kl`,
`s1`) used by `calc_dist_weights()` were chosen, and the visualizations they
were chosen against. It is the counterpart to
`data-raw/martern_parameters.R`, which served the same purpose for the
earlier defaults.

This is a development record, not package documentation — `data-raw/` is
`.Rbuildignore`d, so nothing here ships with the package or is built by
`R CMD check`. It calls package internals directly via `devtools::load_all()`.

## The defaults

```r
gamma = 3e10   # "m3", "m1", "m5", "sq" kernels
kl    = 9e5    # "m3", "m1", "m5", "sq" kernels
s1    = 200    # "bb" kernel
```

These are the values set in `R/calc_distance_weights.R` and
`R/visualize_distance_weights.R`. They were chosen by eye against the
envelope plots below, nudging each until the tapering shape looked right.

**These hyperparameters are not resolution dependent.** They describe our
uncertainty about a biological process — how far a bird's actual path is
likely to stray from the straight line between two cells — and that
uncertainty is a property of the bird, not of the grid we happen to model it
on. So they are fixed literals, deliberately, rather than a formula derived
from `res_m`.

Model resolution enters the kernel in exactly one place, separately from
these three: the nugget (`res_m / 4`), which represents uncertainty about
where within a cell the transition actually started and ended. That part
*is* resolution dependent, and it scales itself.

`res_m = 100000` is used throughout this document only as a reference scale
for reading the plots — a familiar size to think about distances against,
since real BirdFlow models use 100-150 km cells. It is not a scale the
hyperparameters were fit to.

## Checking the shape at a reference scale

```{r}
res_m <- 100000  # reference scale only: typical BirdFlow cell size
cells <- c(1, 2, 5, 10, 15, 30)
lens <- cells * res_m

gamma <- 3e10
kl <- 9e5
s1 <- 200
nugget <- res_m / 4

sd_table <- do.call(rbind, lapply(lens, function(len) {
  data.frame(
    cells = len / res_m,
    len_km = len / 1000,
    m3_sd_end = calc_dist_weights_sd(0, len, res_m, kernel = "m3",
                                     gamma = gamma, kl = kl, s1 = NA),
    m3_sd_mid = calc_dist_weights_sd(len / 2, len, res_m, kernel = "m3",
                                     gamma = gamma, kl = kl, s1 = NA),
    bb_sd_mid = calc_dist_weights_sd(len / 2, len, res_m, kernel = "bb",
                                     gamma = NA, kl = NA, s1 = s1)
  )
}))
sd_table
```

`m3_sd_end` sits exactly at the nugget (`res_m / 4`) for every length — the
bridge conditioning pins the variance to the nugget at the endpoints, so
this is the one column that would move if the model resolution changed. The
`m3_sd_mid` column grows with line length and asymptotes once the line is
long relative to `kl` — toward `sqrt(gamma + nugget^2)` =
`r round(sqrt(gamma + nugget^2) / 1000)` km — but at these `gamma`/`kl`
values that approach is gradual: it's still climbing at a 1500 km line
(~120 km) and doesn't flatten out until well beyond the range shown here.
`bb_sd_mid` keeps growing with `sqrt(length)` (no plateau; that's inherent
to a Brownian bridge with no lengthscale), reaching ~125 km by 1500 km.
Both are biologically plausible: a bird's path straying on the order of
100 km from the straight line over a multi-week migration leg, tapering to
the endpoints and widest in between.

## Envelope plots

The envelope plots show the region within 1.96 standard deviations of the
line — 95 percent of the kernel's density. These are the plots the defaults
were actually tuned against.

```{r, fig.height=6, fig.width=10}
n_cells <- c(1, 2, 5, 10, 15, 50)
line_lengths <- n_cells * res_m

visualize_distance_weights(res_m = res_m, line_lengths = line_lengths,
                           type = "envelope", kernel = "m3",
                           gamma = gamma, kl = kl)
```

```{r, fig.height=4, fig.width=10}
visualize_distance_weights(res_m = res_m, line_lengths = line_lengths,
                           type = "envelope", kernel = "bb", s1 = s1)
```

Both taper from zero at the endpoints and widen toward the middle.

## Weight rasters

The envelope traces the kernel's spread but says nothing about the weight
`calc_dist_weights()` actually returns, which also depends on the detection
radius: the weight is the probability mass falling within `radius_m` of the
point, so it peaks on the line and falls off with the density. The raster
plots below show that weight directly, with `radius_m` at its default of
`res_m / 2`. Fewer and shorter line lengths than above, since each facet is
a 100x100 grid of `calc_dist_weights()` calls.

```{r, fig.height=4, fig.width=10}
raster_lengths <- c(2, 5, 10) * res_m

visualize_distance_weights(res_m = res_m, line_lengths = raster_lengths,
                           radius_m = res_m / 2, type = "raster",
                           kernel = "m3", gamma = gamma, kl = kl)
```

```{r, fig.height=4, fig.width=10}
visualize_distance_weights(res_m = res_m, line_lengths = raster_lengths,
                           radius_m = res_m / 2, type = "raster",
                           kernel = "bb", s1 = s1)
```

Weight is highest near the start and end of each line, where the spread is
pinched to the nugget and so nearly all the density falls inside `radius_m`,
and lowest mid-line where the same density is spread over a much wider band.
That is the intended behavior: a transition is most informative about
locations near its endpoints.

## Sweeping `gamma` at a fixed `kl`

A closer look at how sensitive the `"m3"` envelope is to `gamma`, holding
`kl` at the chosen default (`9e5`): three line lengths as column facets,
`gamma` from `1e10` to `5e10` in `1e10` steps as row facets.

```{r, fig.height=9, fig.width=8}
n_cells_sweep <- c(2, 5, 15)
line_lengths_sweep <- n_cells_sweep * res_m
gamma_values <- seq(1e10, 5e10, by = 1e10)
kl_fixed <- 9e5

length_levels <- paste(line_lengths_sweep / 1000, "km")
gamma_levels <- paste0("gamma = ", format(gamma_values, scientific = TRUE))

sweep <- do.call(rbind, lapply(gamma_values, function(g) {
  do.call(rbind, lapply(line_lengths_sweep, function(len) {
    dist_along <- seq(0, len, length.out = 100)
    sd_m <- calc_dist_weights_sd(dist_along, rep(len, 100), res_m,
                                 kernel = "m3", gamma = g, kl = kl_fixed,
                                 s1 = NA)
    data.frame(
      dist_along_km = dist_along / 1000,
      ymin = -1.96 * sd_m / 1000,
      ymax = 1.96 * sd_m / 1000,
      length_label = factor(paste(len / 1000, "km"), levels = length_levels),
      gamma_label = factor(paste0("gamma = ", format(g, scientific = TRUE)),
                           levels = gamma_levels)
    )
  }))
}))

ggplot(sweep, aes(x = dist_along_km, ymin = ymin, ymax = ymax)) +
  geom_ribbon(fill = "steelblue", alpha = 0.6) +
  facet_grid(gamma_label ~ length_label) +
  coord_fixed() +
  labs(x = "Distance along line (km)",
       y = "Perpendicular distance from line (km)",
       title = "\"m3\" kernel envelope: gamma sweep at kl = 9e5",
       subtitle = paste0("res_m = ", res_m)) +
  scale_y_continuous(breaks = scales::breaks_width(res_m / 1000 * 2)) +
  scale_x_continuous(breaks = scales::breaks_width(res_m / 1000 * 2))
```

Across this range, `gamma` mainly sets how wide the envelope gets once the
line is long enough for the kernel to fully decorrelate (the 1500 km
column), while the 200 and 500 km columns are still partway through the
taper and change more subtly. `kl = 9e5` is what produces that gradual
widen-then-plateau shape across all three lengths, rather than saturating
immediately (small `kl`) or barely moving at all (`kl` much larger than the
longest line).

## Takeaway

- Final defaults, chosen by eye against the envelope plots above:
  `gamma = 3e10`, `kl = 9e5` for the `"m3"`/`"m1"`/`"m5"`/`"sq"` kernels,
  and `s1 = 200` for `"bb"`.
- Applied as fixed literal defaults in `R/calc_distance_weights.R` and
  `R/visualize_distance_weights.R`. Fixed rather than `res_m`-derived
  because they describe biological movement uncertainty, which is
  independent of the model grid; only the nugget (`res_m / 4`) is
  resolution dependent.
- Both kernels give spread on the order of a grid cell, tapering to the
  nugget at the endpoints, with weight concentrated near those endpoints.
