---
title: "BirdFlowR"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{BirdFlowR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "65%",
  fig.width = 6,
  fig.height = 5
)
```

## Setup

### Install packages
```{r install, eval = FALSE }
installed <- rownames(installed.packages())
if (!"remotes" %in% installed)
  install.packages("remotes")
if (!"rnaturalearthdata" %in% installed)
  install.packages("rnaturalearthdata")
remotes::install_github("birdflow-science/BirdFlowModels")
remotes::install_github("birdflow-science/BirdFlowR", build_vignettes = TRUE)
```

### Load libraries
```{r setup}
library(BirdFlowModels)
library(BirdFlowR)
library(terra)
library(sf)
```


### Load model

The BirdFlow Science team has shared a [collection of fitted models](`r paste0(birdflow_options("collection_url"), "index.html")`) for use with
the BirdFlowR package. The website includes reports on each species.  

We can also access the collection index through the package.

```{r load index}
# Load and print index
index <- load_collection_index()
print(index[, c("model", "common_name")])
```

And we can load a model from the collection based on the `model` column from 
the index.  
**Note:** in the vignette this block isn't executed.

```{r load model, eval = FALSE}
# Load a specific model
bf <- load_model("amewoo_prebreeding") # caches locally and loads from cache
```
This loads the smaller example model instead for efficiency of package building
and testing, but do not use this one for science!
```{r load example model, eval = TRUE}
bf <- BirdFlowModels::amewoo # example and test dataset
```

## Access basic information 

`dim()`, `nrow()`, and `ncol()` all report on raster dimensions associated with the model.
`n_active` is the total number of cells that the BirdFlow model can route birds through and is a subset of the cells in the raster.  
`n_transitions()` and `n_distr()` report on temporal dimensions. If the model `is_cyclical()`, they will be equal.

```{r access}
# Methods for base R functions:
dim(bf)
c(nrow(bf), ncol(bf))
bf # same as print(bf)

# BirdFlowR functions
n_active(bf)
n_transitions(bf)
n_timesteps(bf)

# Contents
has_marginals(bf)
has_distr(bf)
has_transitions(bf)
is_cyclical(bf)
```

## Species information and metadata
`species_info()` and `get_metadata()` take a BirdFlow object as their first argument.  An optional second argument allows specifying a specific item, if omitted a list is returned with all the available information.

`species(bf)` is a shortcut for `species_info(bf, "common_name")`

Use `?species_info()` to see descriptions of all the available information.  Dates associated with migration and resident seasons are likely to be useful.

```{r BirdFlow specific info}
species(bf)
species(bf, "scientific")
species_info(bf, "prebreeding_migration_start")
si <-  species_info(bf) # list with all species information
md <- get_metadata(bf)  # list with all metadata
get_metadata(bf, "birdflow_model_date") # date model was exported from python

validate_BirdFlow(bf)  # throws error if there are problems
```
## Spatial aspects

BirdFlow models are based on a raster representation of a time series of species
distributions and contain all the spatial information necessary to recreate 
those distributions and to define how the raster is positioned in space.
BirdFlowR uses the **terra** package to import raster data and provides BirdFlow 
methods for functions defined in the terra package - so that you can use those 
functions on BirdFlow objects.

`crs()` returns the coordinate reference system - useful if you need to project other data to match the BirdFlow object.
`res()`, `xres()`, and `yres()` describe the dimensions of individual cells in the model.
`ext()` returns a terra extent object.  
`compare_geom()` tests if the extent, resolution, and CRS of two objects is the
same. BirdFlowR includes methods to compare BirdFlow models with each other and
with terra objects.

```{r spatial aspects terra}
# Methods for terra functions:
a <- crs(bf) # well known text (long)
crs(bf, proj = TRUE)  # proj4 string
res(bf)
c(xres(bf), yres(bf)) # same as res(bf)
ext(bf)
c(xmin(bf), xmax(bf), ymin(bf), ymax(bf)) # same as ext(bf)

# Compare geometries - do they have the same CRS, extent, and cell size
compareGeom(bf, rast(bf))
```

BirdFlow objects also play nicely with the *sf* package. 

```{r sf}
bb <- sf::st_bbox(bf)
crs <- sf::st_crs(bf)
```

## Retrieve and plot distributions

A distribution in BirdFlow is stored as a vector of values that correspond to only the active cells (`n_active()`) in the model.  Multiple distributions are stored as matrices with `n_active()` rows and a column for each distribution.

We can retrieve distributions in this format with `get_distr()`.
Use timestep, character dates, date objects, or "all" to specify which distributions to retrieve.

Retrieve the first distribution and compare its length to the number of active cells.
```{r single distribution}
d <- get_distr(bf, 1) # get first timestep distribution
length(d)  # 1 distribution so d is a vector
n_active(bf)  # its length is the the number of active cells in the model
```

Get 5 distributions, the result is a matrix in which each column is a  distribution with a row for each active cell. 
```{r multiple distributions}
d <- get_distr(bf, 26:30)
dim(d)
head(d, 3)
```

We can also specify distributions with dates, or use "all" to retrieve all the distributions. 
```{r get_distr options}
d <- get_distr(bf, "2022-12-15") # from character date
d <- get_distr(bf, "all")  # all distributions
d <- get_distr(bf, Sys.Date())  # Using a Date object
```

Use `rasterize_distr()` to convert a distribution to a SpatRaster defined in the terra package. The second argument, the BirdFlow 
model, is needed for the spatial information it contains. 
```{r get distributions}
d <- get_distr(bf, c(1, 26)) # winter and summer
r <- rasterize_distr(d, bf) # convert to SpatRaster
```

Alternatively convert directly from BirdFlow to SpatRaster with `rast()`. The second (optional) argument `which` accepts the same inputs as `which` in `get_distr()`.

```{r rast, fig.width=8, fig.height=4, out.width='100%'}
r <- rast(bf) # all distributions
r <- rast(bf, c(1, 26))  # 1st, and 26th timesteps.
plot(r)
```

BirdFlowR provides convenience wrappers to functions in **rnaturalearth**
that load vector data and then crop and transform it to make it suitable for plotting with BirdFlow output.  

***Note:** until **rnaturalearth** has fully transitioned
away from legacy packages you may see a warning about them, but **BirdFlowR** 
does not use the legacy packages or data formats itself.*

```{r }
r <- rast(bf, species_info(bf, "prebreeding_migration_start"))
plot(r)
coast <- get_coastline(bf)  # lines
plot(coast, add = TRUE)
```

# Forecasting
In this section we will sample a single starting location from the winter
distribution and project it forward. We will generate a distribution of 
predicted breeding grounds for birds that wintered at the starting location.

Set predict parameters.
```{r predict parameters}
    start <- 1     #  winter
    end <-  26     # summer
```

## Sample starting distribution
`sample_distr()` will sample from one or more input distribution to select a 
single location per distribution. The result is one or more distributions 
with ones in the selected location(s) and zero elsewhere.  

```{r starting location}
set.seed(0)
d <- get_distr(bf, start)
location <- sample_distr(d)

print(i_to_xy(which(as.logical(location)), bf))  # starting coordinates

```

## Project forward from this location to summer
`predict()` returns the distribution over time as a matrix with
one column per timestep. 

The plot shows where birds that winter at a particular location are
likely to be as the year progresses and ultimately where they might spend their
summer. The probability density spreads as the weeks progress.
```{r predict, out.width='100%'}
f <- predict(bf, distr = location, start = start, end = end,
             direction = "forward")

r <- rasterize_distr(f[, c(1, 7, 14, 19)], bf)
plot(r)
```

Additionally, we can calculate the difference between the projected distribution
and the distribution of the species as a whole at the same timestep. 
```{r probability over time}
projected <- f[, ncol(f)]  # last projected distribution
diff <-  projected - get_distr(bf, end)
plot(rasterize_distr(diff, bf))
```

# Generate synthetic routes 
Here we sample locations from the American Woodcock winter 
distribution and generate routes to their summer grounds.

Set route parameters.
```{r route parameters}
n_positions <-  15 # number of starting positions
start <- 1         # starting timestep (winter)
end <- 26          # ending timestep (summer)
```

## Generate starting locations 
First extract the winter distribution, then use `sample_locations()` with 
`n = n_positions` to sample the input distribution repeatedly. The result is a
matrix in which each column has a single '1' representing the sampled location.
```{r starting locations}
d <- get_distr(bf, start)
locations  <- sample_distr(d, n = n_positions, bf = bf, format = "xy")
x <- locations$x
y <- locations$y
```

Plot the starting (winter) distribution and sampled locations.
```{r plot starting distribution}
winter <- rasterize_distr(d, bf)
plot(winter)
points(x, y)
```

## Generate routes
`route()` will generate synthetic routes for each starting position.
`route()` returns a BirdFlowRoutes object which has a `data` feature for data frame 
with a row for each timestep of each route, but also includes some additional
spatial and temporal information from the BirdFlow object.

```{r route, fig.show='hide'}
rts <- route(bf, x_coord = x, y_coord = y, start = start, end = end)
head(rts$data, 4)
```

The `route()` function can sample starting locations from the distribution for
the starting timestep so the following is equivalent to the preceding two sections.

```{r route with sample}
rts2 <- route(bf,  n = n_positions,  start = start, end = end)

```

We can specify the date range with any arguments supported by
`lookup_timestep_sequence()` so an alternative to the above with slightly
different start and end dates is to use the season argument.  Here we route
during the prebreeding migration. 

```{r route with season}
rts3 <- route(bf, n = n_positions, season = "prebreeding")
```

## Using base R plotting to plot routes 
Plot the route lines over the summer distribution along with points at the 
starting and ending positions. 
```{r plot routes with base R}
d <- get_distr(bf, end)
summer <- rasterize_distr(d, bf)

line_col <- rgb(0, 0, 0, .2)
pt_col <- rgb(0, 0, 0, .5)

plot(summer)
points(x, y, cex = .4, col = pt_col, pch = 16) # starting points

rts_sf <- sf::st_as_sf(rts$data, coords = c("x", "y"), crs = rts$geom$crs)
lines_sf <- rts_sf |>
  dplyr::group_by(route_id) |>
  dplyr::summarize(geometry = st_combine(geometry)) |>
  sf::st_cast("LINESTRING")

plot(lines_sf, add = TRUE, col = line_col)  # routes

end_pts <- rts$data[rts$data$timestep == end, ]  # end points
points(x = end_pts$x, y = end_pts$y,
       cex = 0.4, pch = 12, col = pt_col)

title(main = species(bf))
```

## Or use plot_routes()

`plot_routes()` will visualize time as a color gradient using **ggplot2** with
stop point dots that indicate how long a bird was at each stopping point.

```{r plot routes}
plot_routes(rts, bf)
```
