---
title: "BirdFlowR introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BirdFlowR introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "65%",
  fig.width = 6,
  fig.height = 5
)
```
A brief introduction to core BirdFlowR functions. 

## Install packages and load BirdFlow model

Install packages
```{r install, eval = FALSE }
installed <- rownames(installed.packages())
if(!"devtools" %in% installed)
  install.packages("devtools")
if(!"BirdFlowR" %in% installed)
  devtools::install_github("birdflow-science/BirdFlowR")
```

Currently the example model isn't part of this or any other package so to run
the code in this vignette you will need to download the model file and edit 
the model_path line below to point to the local file.

[Download](https://drive.google.com/file/d/1KW0jy44_PbAI6askkVZn9xP1_bBz-HlC/edit)
(634 MB from google drive)

```{r setup}
library(BirdFlowR)
library(terra)

model_path <- "../../Models/amewoo_sparse.Rds"  # This will need to be changed
bf <- readRDS(model_path)

```

## Access basic information
```{r access}
# Methods for base R functions:
dim(bf)
nrow(bf)
ncol(bf)

# New functions:
n_active(bf)
n_transitions(bf)
n_timesteps(bf)
species(bf)  
species(bf, "scientific")
# species(bf, "all") # list with all species metadata


# Methods for terra functions:
a <- crs(bf) # well known text (long)
crs(bf, proj = TRUE)  # proj4 string
res(bf)
xres(bf)
yres(bf)
ext(bf)
c(xmin(bf), xmax(bf), ymin(bf), ymax(bf))


# Print method for BirdFlow objects:
print(bf)  
```

## Retrieve and plot distributions

Use timestep, character dates, date objects, or "all" to specify which 
distributions to retrieve.

Retrieve the first distribution and compare its length to the number of active 
cells.
```{r single distribution}
d <- get_distr(1, bf) # get first timestep distribution 
length(d)  # 1 distribution so d is a vector
n_active(bf)  # its length is the the number of active cells in the model
```

Get 5 distributions, the result is a matrix in which each column is a 
distribution with a row for each active cell. 
```{r multiple distributions}
d <- get_distr(26:30, bf) 
dim(d)
head(d, 3)
```
We can also specify distributions with dates, or use "all" to retrieve all the distributions. 
```{r get_distr options}
d <- get_distr("2022-12-15", bf) # from character date
d <- get_distr("all", bf)  # all distributions
d <- get_distr(Sys.Date(), bf)  # Using a Date object 
```

Use `rasterize_distr()` or `rast()`to convert to SpatRaster defined in the terra package. 
With `rasterize_distr()` the second argument is the BirdFlow model.
```{r get distributions}
d <- get_distr(c(1, 26), bf) # winter and summer
r <- rasterize_distr(d, bf) # convert to SpatRaster 
```
Alternatively convert directly from BirdFlow to SpatRast with `rast()` the 
second, optional, argument `which` takes the same input as `get_distr()`.

```{r rast, fig.width=8, fig.height=4, out.width='100%'}
r <- rast(bf) # all distributions
r <- rast(bf, c(1, 26))
plot(r)
```

# Forecasting
In this section we will sample a single starting location from the winter distribution and project it forward to generate a distribution of predicted breeding grounds for birds that wintered at the starting location.

Set forecasting parameters
```{r forecast parameters}
start <- 1         # starting timestep (winter)
end <- 26          # ending timestep (summer)
```

## Sample starting distribution
`sample_distr()` will probabilistically assign a 1 to a single location per input distribution and set the remaining cells to 0.  
```{r starting location}
set.seed(0)
d <- get_distr(start, bf)
location <- sample_distr(d)
```

## Project forward from this location to summer
Forecast returns the distribution over time as a matrix with
one column per timestep. 

The plot shows how the probability density spreads over time from the starting 
location for a subset of the timesteps.  The last panel represents the summer distribution.
```{r forecast, out.width='100%'}
f <- forecast(bf, location, start, end, "forward")
r <- rasterize_distr(f[, c(1, 9, 18, 26)], bf)
plot(r)
```
Alternatively we can calculate the difference between the projected distribution and the distribution of the species as a whole at the same timestep.
```{r probability over time}
projected <- f[ , ncol(f)]  # last projected distribution
diff <-  projected - get_distr(end, bf) 
plot(rasterize_distr(diff, bf))
```

# Generate synthetic routes 
In this section we sample locations from the American Woodcock winter 
distribution and then generate routes to their summer grounds.

Set route parameters
```{r route parameters}
n_positions <-  15 # number of starting positions
n_each <- 1        # how many birds to start at each
start <- 1         # starting timestep (winter)
end <- 26          # ending timestep (summer)
```

## Generate starting locations 
First extract the winter distribution, then use `sample_locations()` with `n = n_positions` to sample the input distribution repeatedly resulting in a matrix in which each column has a single 1 representing the sampled location.
```{r starting locations}
d <- get_distr(start, bf)
locations  <- sample_distr(d, n = n_positions)  
```
Collapse the locations down to a vector of the index of each
non-zero value and then convert to x and y coordinates.
```{r starting locations as row and column}
ind <- apply(locations, 2, function(x) which( as.logical(x) ) )
x <- i_to_x(ind, bf)
y <- i_to_y(ind, bf)
```
Plot the starting (winter) distribution and sampled locations
```{r plot starting distribution}
winter <- rasterize_distr(d, bf)
plot(winter)
points(x, y)
```

## Generate routes
`route()` will generate synthetic routes for each starting position.
Currently `route()` returns a list with two items (this may change):

* `points` a data.frame with a row for each timestep of each route
* `lines` sf object containing a line for each route
```{r route, fig.show='hide'}
rts <- route(bf, x_coord = x, y_coord = y, start = start,
             end = end, n = n_each)
head(rts$points, 4)
```

## Plot routes 
Plot the route lines over the summer distribution along with points at the starting and ending positions.
```{r plot routes}
d <- get_distr(end, bf)
summer <- rasterize_distr(d, bf)

line_col <- rgb(0, 0, 0, .2)
pt_col <- rgb(0, 0, 0, .5 )

plot(summer) 
points( x, y, cex = .4, col = pt_col, pch = 16) # starting points
plot(rts$lines, add = TRUE, col = line_col)  # routes
end_pts <- rts$points[rts$points$timestep == end, ]  # end points
points(x = end_pts$x, y = end_pts$y,
       cex = 0.4, pch = 12, col = pt_col )

title(main = bf$metadata$species)

```
