---
title: "BirdFlowR introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BirdFlowR introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "65%",
  fig.width = 6,
  fig.height = 5
)
```
A brief introduction to BirdFlowR.

## Setup

Install packages
```{r install, eval = FALSE }
installed <- rownames(installed.packages())
if(!"devtools" %in% installed)
  install.packages("devtools")
if(!"BirdFlowR" %in% installed)
  devtools::install_github("birdflow-science/BirdFlowR")
if(!"BirdFlowModels" %in% installed)
  devtools::install_github("birdflow-science/BirdFlowModels")
```

Load libraries and define local BirdFlow model (`bf`)
```{r setup}
library(BirdFlowR)
library(BirdFlowModels)
library(terra)

bf <- amewoo  # from BirdFlowModels

```

## Access basic information
```{r access}
# Methods for base R functions:
dim(bf)
c( nrow(bf), ncol(bf) )

# New functions:
n_active(bf)
n_transitions(bf)
n_timesteps(bf)

species(bf)  
species(bf, "scientific")
# species(bf, "all") # list with all species metadata


# Methods for terra functions:
a <- crs(bf) # well known text (long)
crs(bf, proj = TRUE)  # proj4 string
res(bf)
c( xres(bf), yres(bf) ) # same as res(bf)
ext(bf)
c(xmin(bf), xmax(bf), ymin(bf), ymax(bf)) # same as ext(bf)

# Print method for BirdFlow objects:
print(bf)  
```

## Retrieve and plot distributions

Use timestep, character dates, date objects, or "all" to specify which 
distributions to retrieve.

Retrieve the first distribution and compare its length to the number of active 
cells.
```{r single distribution}
d <- get_distr(1, bf) # get first timestep distribution 
length(d)  # 1 distribution so d is a vector
n_active(bf)  # its length is the the number of active cells in the model
```

Get 5 distributions, the result is a matrix in which each column is a 
distribution with a row for each active cell. 
```{r multiple distributions}
d <- get_distr(26:30, bf) 
dim(d)
head(d, 3)
```

We can also specify distributions with dates, or use "all" to retrieve all the distributions. 
```{r get_distr options}
d <- get_distr("2022-12-15", bf) # from character date
d <- get_distr("all", bf)  # all distributions
d <- get_distr(Sys.Date(), bf)  # Using a Date object 
```

Use `rasterize_distr()` to convert a distribution to a SpatRaster defined in the
terra package. The second argument, the BirdFlow 
model, is needed for the spatial information it contains. 
```{r get distributions}
d <- get_distr(c(1, 26), bf) # winter and summer
r <- rasterize_distr(d, bf) # convert to SpatRaster 
```

Alternatively convert directly from BirdFlow to SpatRast with `rast()` the 
second, optional, argument `which` accepts the same inputs a as `which` in
`get_distr()`.

```{r rast, fig.width=8, fig.height=4, out.width='100%'}
r <- rast(bf) # all distributions
r <- rast(bf, c(1, 26))  # 1st, and 26th timesteps.
plot(r)
```

# Forecasting
In this section we will sample a single starting location from the winter distribution and project it forward to generate a distribution of predicted breeding grounds for birds that wintered at the starting location.

Set forecast parameters.
```{r forecast parameters}
start <- 1         # starting timestep (winter)
end <- 26          # ending timestep (summer)
```

## Sample starting distribution
`sample_distr()` will sample from one or more input distribution to select a 
single location per distribution. The result is one or moe distributions 
with ones in the selected location(s) and zero elsewhere.  
```{r starting location}
set.seed(0)
d <- get_distr(start, bf)
location <- sample_distr(d)


print( i_to_xy( which( as.logical(location) ), bf) )  # starting coordinates

```

## Project forward from this location to summer
Forecast returns the distribution over time as a matrix with
one column per timestep. 

The plot shows where birds that winter at a particular location are
likely to be as the year progresses and ultimately where they might spend their
summer. The probability density spreads as the weeks progress.
```{r forecast, out.width='100%'}
f <- forecast(bf, location, start, end, "forward")
r <- rasterize_distr(f[, c(1, 9, 18, 26)], bf)
plot(r)
```

Alternatively we can calculate the difference between the projected distribution
for bird's wintering at that particular location and the distribution of the 
species as a whole at the same timestep.
```{r probability over time}
projected <- f[ , ncol(f)]  # last projected distribution
diff <-  projected - get_distr(end, bf) 
plot(rasterize_distr(diff, bf))
```


# Generate synthetic routes 
In this section we sample locations from the American Woodcock winter 
distribution and then generate routes to their summer grounds.

Set route parameters.
```{r route parameters}
n_positions <-  15 # number of starting positions
n_each <- 1        # how many birds to start at each
start <- 1         # starting timestep (winter)
end <- 26          # ending timestep (summer)
```

## Generate starting locations 
First extract the winter distribution, then use `sample_locations()` with `n = n_positions` to sample the input distribution repeatedly resulting in a matrix in which each column has a single 1 representing the sampled location.
```{r starting locations}
d <- get_distr(start, bf)
locations  <- sample_distr(d, n = n_positions)  
```

Collapse the locations down to a vector of the index of each
non-zero value and then convert to x and y coordinates.
```{r starting locations as row and column}
ind <- apply(locations, 2, function(x) which( as.logical(x) ) )
x <- i_to_x(ind, bf)
y <- i_to_y(ind, bf)
```

Plot the starting (winter) distribution and sampled locations
```{r plot starting distribution}
winter <- rasterize_distr(d, bf)
plot(winter)
points(x, y)
```

## Generate routes
`route()` will generate synthetic routes for each starting position.
Currently `route()` returns a list with two items (this may change):

* `points` a data.frame with a row for each timestep of each route
* `lines` sf object containing a line for each route
```{r route, fig.show='hide'}
rts <- route(bf, x_coord = x, y_coord = y, start = start,
             end = end, n = n_each)
head(rts$points, 4)
```

## Plot routes 
Plot the route lines over the summer distribution along with points at the starting and ending positions.
```{r plot routes}
d <- get_distr(end, bf)
summer <- rasterize_distr(d, bf)

line_col <- rgb(0, 0, 0, .2)
pt_col <- rgb(0, 0, 0, .5 )

plot(summer) 
points( x, y, cex = .4, col = pt_col, pch = 16) # starting points
plot(rts$lines, add = TRUE, col = line_col)  # routes
end_pts <- rts$points[rts$points$timestep == end, ]  # end points
points(x = end_pts$x, y = end_pts$y,
       cex = 0.4, pch = 12, col = pt_col )

title(main = bf$metadata$species)

```
