---
title: "Calculate movement among polygons"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Calculate movement among polygons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>",
  out.width = "65%",
  fig.width = 6,
  fig.height = 5
)

library(BirdFlowR)
library(terra)
```

## Overview

The goal here is to estimate movement among polygons over a defined period of
time using a BirdFlow model. The result is square movement matrix where 
the rows correspond to starting polygons and the columns to ending polygons.
Values are the proportion of the total population making each movement, which  
can be scaled to represent numbers of birds. Optionally each value can 
be adjusted by the area of the two associated polygons to get a mean movement
per square km.

The initial use case was to create a predictor for a model of 
avian influenza spread. In the example here the polygons are US States but
the variable (`polys`) and the sole required column name (`id`) are kept
deliberately generic so the code will make sense with any set of polygons.

This will likely be turned into a function but the vignette should 
remain a useful demonstration of how to work with BirdFlow models.

### The approach:

1. Load model, set start and end times for connectivity window, and set
total population.
1. Download and prepare polygons
1. Calculate the overlap between active cells in the BirdFlow model and
the polygons.
1. Generate starting distributions for each polygon that represent the
portion of the species distribution that is in each polygon at the start
of the movement.
1. Run `predict()` using distributions and start and end time. 
1. Drop time dimension from result preserving just the end distributions.
1. Convert the ending distribution from model cells to polygons. 
1. Optionally, multiply by the species' total population (within the Americas) 
   to convert from proportion of population to estimated number of birds
1. Optionally, divide by the area of both the starting and ending state to 
 make the estimates per unit area - eliminating the polygon size as a factor 
 in the connectivity. This step is not appropriate for all use cases.

## Load model and set parameters

This example uses the American Woodcock dataset from 
**BirdFlowModels** which is small and runs fast but isn't fully vetted for scientific use. 

Use `load_model()` for real work. 

```{r load BirdFlow model, set parameters}
bf <- BirdFlowModels::amewoo  # you will likely use load_model() instead
start <- BirdFlowR::lookup_timestep("2024-01-15", bf)
end <- BirdFlowR::lookup_timestep("2024-03-20", bf)
population <- 3500000
```


### Visualize the starting distribution from the model

```{r plot starting distribution}
plot_distr(get_distr(bf, start), bf)
```

## Load Polygons

Use `get_states()` to get the state boundaries within  BirdFlow
object extent from the[Natural Earth](https://www.naturalearthdata.com/) 
data set. 

If you have a polygon shapefile already you could use 
`polys <- sf::read_sf(shapefile_path)` 
and skip this section.

```{r rnaturalearth, warning = FALSE} 
# In March of 2025 downloading from the census became less reliable.
# This section is an alternative approach which uses get_states()


polys <- get_states(bf, country = "United States of America", 
                    keep_buffer = TRUE, keep_attributes = TRUE)

polys <- dplyr::select(polys, NAME = gn_name)


```

## Prepare polygons for analysis

* Add an `id` field; in this example it is the state names.
* Transform to match the BirdFlow Coordinate Reference System (CRS).
* Calculate the area in square kilometers of each polygon (`sq_km`).
* Add centroid coordinates (`x`, `y`) in the BirdFlow CRS.

```{r prepare polygons}

# Add generic "id" column (use state name) 
polys$id <- polys$NAME

# Order by id (optional)
polys <- polys[order(polys$id), ]

# Transform to match the BirdFlow model CRS 
polys <- sf::st_transform(polys, crs(bf))

# add sq_km to polygons 
# This assumes the projection is in meters which is true for BirdFlow models
polys$sq_km <- sf::st_area(polys)  / 1000^2  |> as.vector()

# Add centroids (only used for the plotting at end of document)
centroids <- sf::st_centroid(polys[, "geometry"]) |> sf::st_coordinates()
colnames(centroids) <- c("x", "y")
polys <- cbind(polys, centroids)

```

## Calculate overlap between active BirdFlow cells and polygons

Make the `overlap` matrix.  Rows correspond to active cells, and columns 
to the external polygons. The values are the proportion of the cell that 
overlaps the polygon. 

```{r calculate overlap, echo = TRUE}

# Convert active cells to polygons
cell_polys <- rasterize_distr(seq_len(n_active(bf)), bf, format = "terra") |>
 terra::as.polygons() |> 
  sf::st_as_sf()
names(cell_polys)[1] <- "i" # BirdFlow location index "i"

# Intersect cells with polygons
suppressWarnings( # Attribute values assumed to be spatially constant... 
  intersection_polys <- sf::st_intersection(cell_polys, polys) 
)
intersection_polys <- intersection_polys[ , c("i", "id")]
intersection_polys$area <- sf::st_area(intersection_polys)
sq_m_per_cell <- prod(res(bf)) 
intersection_polys$prop_of_cell <- intersection_polys$area / sq_m_per_cell

# Calculate the proportion of each cell that overlaps each polygon
# Rows = active cells, columns = polygons 
overlap <- matrix(0, nrow = n_active(bf), ncol = nrow(polys), 
                  dimnames = list(i = seq_len(n_active(bf)),
                                  id = polys$id))
intersection_polys$row <- match(intersection_polys$i, seq_len(n_active(bf)))
intersection_polys$col <- match(intersection_polys$id, polys$id)                  
row_col_index <- intersection_polys[ , c("row", "col")] |> 
  sf::st_drop_geometry() |> 
  as.matrix()
overlap[row_col_index] <- intersection_polys$prop_of_cell |> as.vector()

```

## Build initial distributions

This essentially clips the initial distribution separately to each polygon while 
assigning values from boundary cells in proportion to their overlap with the
polygon.

Note, a single BirdFlowR distribution is a vector of length `n_active(bf)` 
multiple can be stored in a matrix with `n_active(bf)` rows.  The matrix here
has a column (distribution) for each polygon in `polys`.
 
```{r }

start_distr <- matrix(get_distr(bf, start), 
                              nrow = n_active(bf), 
                              ncol = nrow(polys), 
                              byrow = FALSE, 
                              dimnames = list(i = NULL, 
                             start_poly = polys$id)) 

start_distr <- start_distr * overlap

```

```{r select largest, echo = FALSE, eval = TRUE}
# Doing this ahead of time so I can pull the name in text below.
sel <- which.max(apply(start_distr, 2, sum))
```

### Visualize the distribution for one polygon 

Visualize the initial distribution 
for the polygon with the highest initial abundance (`r polys$id[sel]`).

```{r visualize largest distribution, warning = FALSE, message = FALSE}
sel <- which.max(apply(start_distr, 2, sum))
plot_distr(start_distr, bf, subset = sel) + 
  ggplot2::geom_sf(data = polys, 
                   inherit.aes = FALSE, 
                   linewidth = 0.2, 
                   color = "black", 
                   fill = NA)
```

## Project the distributions forward

Use `BirdFlowR::predict()` to project the distributions forward.
With multiple input distributions the output will be a three dimensional array. 
Select the last element of the third dimension to get just the ending 
distributions in a matrix.
The values in these distributions represent the proportion of the population 
that started in the corresponding polygon (columns) that ended in each
active cell (rows).

```{r predict}

# Project distributions forward to end date
pred <- predict(bf, distr = start_distr, start = start, end = end)

# Just keep last distribution 
end_distr <- pred[ , , dim(pred)[3]] 

# Preserve dimension names
dimnames(end_distr) <- dimnames(start_distr) 

```

### Visualize the ending distribution 

This shows the ending distribution for the birds that started in 
`r polys$id[sel]`.

```{r visualize ending distribution, message = FALSE, warning = FALSE} 
plot_distr(end_distr, bf, sel) +
  ggplot2::geom_sf(data = polys, 
                   inherit.aes = FALSE, 
                   linewidth = 0.2, 
                   color = "black", 
                   fill = NA)
```


## Collapse the ending cells into ending polygons

Convert the `end_distr` destination cells into
destination polygons based on the overlap proportions in `overlap`.

The first "slow" way is perhaps easier to understand. The faster way produces
the same result with matrix multiplication.
Either way `move` is a matrix with rows for the starting polygons,
columns for the destination polygons,
and  values that are the proportion of the population
making the transition between the two.

```{r destination states}
slow <- FALSE
if (slow) {
  # Here we loop through each cell of the movement matrix filling in values
  # one by one.
  n <- nrow(polys)
  move <- matrix(NA, n, n, dimnames = list(from = polys$id, to = polys$id))
  for(from_i in seq_len(n)) for (to_i in seq_len(n)) {
    # Rows are source polygons, columns for destination polygons. 
    # Values are the proportion of the total population making the transition
    # between the corresponding polygons.
    move[from_i, to_i] <- sum(end_distr[, from_i] * overlap[, to_i])
  } 
} else {
  # Use matrix multiplication to do the same thing faster
  move <- t(end_distr) %*% overlap
  dimnames(move) <- list(from = polys$id, to = polys$id)
}

```

### Visualize the polygonized end distribution

Plot the proportion of the total population that moved from  
`r polys$id[sel]` to each each state.

```{r visualize polygon distr}
polys$Proportion  <- move[sel, ]

gradient_colors <-
      c("#EDDEA5", "#FCCE25", "#FBA238", "#EE7B51", "#DA596A", "#BF3984",
        "#9D189D", "#7401A8", "#48039F", "#0D0887")

ggplot2::ggplot(data = polys) + 
  ggplot2::geom_sf(ggplot2::aes(fill = Proportion)) + 
   ggplot2::scale_fill_gradientn(colors = gradient_colors) + 
  ggplot2::ggtitle(paste0(species(bf), " movement from ", polys$id[sel]), 
                   subtitle = paste0(lookup_date(start, bf), " to ", 
                                     lookup_date(end, bf)))

```

## Adjust for population (optional)

You can multiply the movement by the total population of the species to 
estimated the number of individuals making each movement. This is likely 
necessary to aggregate or compare output across species. 

```{r adjust for population}
# Convert to absolute numbers of Birds expected to move between 
# each pair of polygons
move_count  <- move * population
```

## Adjust for area (optional)

Adjusting for area might be appropriate depending on the application.
If the goal is to capture the total magnitude of movement between each source
and destination polygon than do not adjust for area. 

However, in our initial use case our points were in the specified states but
did not represent the entire state so failing to adjust for state area would
introduce a bias - essentially inflating movement for points that fell in bigger
states. Similarly if you want an average per unit area movement
between states that isn't biased by state area you would adjust for area.

The adjustment is done by dividing the number of individuals moving between the
two polygons by the area of each polygon in sq km to produce the expecting number 
to move between them per sq km at each end. Clearly, though this does 
not capture the fine scale variability within them some of which is captured in 
the original BirdFlow cells. 

```{r adjust for population and area}

# Adjust for area if both the source and destination polgons
n <- nrow(polys)
from_area <- matrix(polys$sq_km, n, n, byrow = FALSE)
to_area <- matrix(polys$sq_km, n, n, byrow = TRUE)
move_area <- move_count / from_area / to_area

```

## Visualize movement

```{r set plot threshold}
plot_count_threshold <- 1000 # only plot lines for big movements
```

Plot non-zero estimated counts of birds moving between polygons and
indicate magnitude with line thickness. Lines are only shown for movements of 
at least `r plot_count_threshold` individuals.

```{r plot connections}
n <- nrow(polys)
long <- data.frame(from =  rep(polys$id, times = n), 
                   to = rep(polys$id, each = n), 
                   count = as.vector(move_count))

# Add coordinates of centroids of to and from polygons
f_mv <- match(long$from, polys$id)
long$f_x <- polys$x[f_mv]
long$f_y <- polys$y[f_mv]
t_mv <- match(long$to, polys$id)
long$t_x <- polys$x[t_mv]
long$t_y <- polys$y[t_mv]

long <- long[long$count >= plot_count_threshold, ]

ggplot2::ggplot(long) + 
  ggplot2::geom_segment(ggplot2::aes(x = f_x, y = f_y, 
                                     xend = t_x, yend = t_y, 
                                     linewidth = count),
                        color = rgb(0, 0, 0, .15)) + 
  ggplot2::scale_linewidth(range = c(0.2, 4)) + 
  ggplot2::theme(axis.title = ggplot2::element_blank(),
                 strip.background = ggplot2::element_blank()) +
  ggplot2::geom_sf(data = polys, 
                   inherit.aes = FALSE, 
                   linewidth = 0.2, 
                   color = "black", 
                   fill = NA)
```
